---
title: "Where pay transparency bites"
subtitle: "Europe's published gender pay gap understates what employers will find inside their own workforces"
author: "Domenico Perroni"
date: 2026-09-17
date-format: "D MMMM YYYY"
format:
  html:
    minimal: true
    embed-resources: true
    css: style.css
    toc: false
    lang: en
    fig-width: 6.8
    fig-height: 4.3
bibliography: ../paper/references.bib
link-citations: true
---

```{r setup}
suppressPackageStartupMessages({library(dplyr); library(tidyr); library(ggplot2)})
for (f in list.files("R", pattern = "[.]R$", full.names = TRUE)) source(f)
a <- run_analysis()
f <- facts(a)
```

::: {.lede}
From June 2027, every employer in the EU with 250 or more workers has to publish its gender pay
gap every year, and to justify, fix or jointly assess any gap of 5% or more between women and men
doing work of equal value. The number most people will benchmark against is the national gap
Eurostat publishes. **It is the wrong benchmark.** Inside the same sector, the gap is wider in
`r f$within_above` of `r f$countries` Member States, and it is wider still in the large employers
that report first.
:::

*Explore every country in the [interactive dashboard](dashboard/) · read the
[working paper](paper.pdf) (PDF)*

## The short version

- **Half of the workforce is in the first wave.** `r f$eu_250` of employees in enterprises with
  10 or more staff work for one with 250 or more, the size that reports every year from
  7 June 2027. Another `r f$eu_50` work in enterprises of 50 to 249.
- **The published EU gap was `r f$eu_2022` in 2022; inside the same sector it was
  `r f$eu_within`.** Women are over-represented in sectors that pay above average, such as
  education and health, and that flatters the national figure.
- **The lowest headline gaps hide the most.** `r f$it$name`: `r f$it$gap` published,
  `r f$it$within` within sectors. `r f$pl$name`: `r f$pl$gap` against `r f$pl$within`.
  `r f$ro$name`: `r f$ro$gap` against `r f$ro$within`.
- **The gap is wider where reporting is annual.** In all `r f$size_countries` countries where
  Eurostat publishes earnings by enterprise size, the gap is larger in enterprises of 250 or more.
  In Poland it is `r f$pl_large` against `r f$pl_small` in smaller ones.
- **What employers will report looks like the inside view, not the headline.** The Directive
  compares women and men in the same organisation and in the same category of workers. Expect
  those numbers to surprise, most of all where the national gap looks small.

## 1. A gap that has been closing

```{r trend, fig.height = 3.6}
fig_trend(a)
```

The unadjusted gap for the EU-27 fell from `r f$eu_2010` in 2010 to `r f$eu_latest` in
`r f$year`. It narrowed in `r f$narrowed` of `r f$compared` countries and widened in
`r f$widened`. In `r f$year` it ranged from `r f$lowest_gap` in `r f$lowest` to
`r f$highest_gap` in `r f$highest`.

That is the figure behind most headlines, and it is computed well: mean gross hourly earnings of
men and women in enterprises with ten or more employees, across the whole economy except public
administration and defence [@eurostat_gpg_meta]. What it cannot show is where the gap sits.

## 2. The headline hides the gap inside sectors

A national gap mixes two things: **where** women and men work, and **how** they are paid once
they are there. Splitting the 2022 gap across the 17 economic sections of the Structure of
Earnings Survey separates the two. The split is symmetric, so the two parts add up to the
published gap exactly and do not depend on taking men's or women's pay as the yardstick.

```{r hidden, fig.height = 6.4}
fig_hidden(a)
```

In `r f$between_negative` countries the "where" part is negative: women are concentrated in
sectors that pay more per hour than the average. In Italy, `r f$it_education` of women in the
survey work in education against `r f$it_education_m` of men, and `r f$it_health` in health and
social work against `r f$it_health_m`. That lowers the Italian headline by `r sub("-", "", f$it$between)`,
and leaves a gap of `r f$it$within` between women and men working in the same sector.

The countries where the headline hides the most are `r f$hidden_top`. Part of the explanation is
well documented: where fewer women work, those who do are often the better paid, which keeps the
average gap low [@olivetti2008]. For an employer, the reason matters less than the consequence.
The Directive looks inside each employer and compares women and men doing the same work or work
of equal value. That comparison is far closer to the within-sector view than to the national
average, and there the gap is wider than the headline suggests.

The result does not hinge on the method. Taking men's pay as the reference, the within-sector
gap exceeds the published one in `r f$within_above_men` countries; taking women's, in
`r f$within_above_women`.

## 3. Finer cells: the gap persists

Sectors are broad. Where Eurostat publishes complete figures by occupation
(`r f$finer_occ` countries) or by sector and occupation together (`r f$finer_so` countries),
the gap between women and men in the same cell stays between `r f$finer_min` and
`r f$finer_max`:

```{r finer}
finer <- a$decomposition |>
  filter(cells != "sector", usable, geo %in% EU27) |>
  mutate(Country = unname(a$names[geo])) |>
  select(Country, cells, official, within) |>
  pivot_wider(names_from = cells, values_from = within) |>
  arrange(Country)
knitr::kable(
  finer |> mutate(across(where(is.numeric), \(x) ifelse(is.na(x), "", fmt_pct(x)))),
  col.names = c("Country", "Published gap", "Within occupation", "Within sector and occupation"),
  align = "lrrr"
)
```

Where the published gap is small, finer cells confirm a wide gap between women and men doing
similar work. Where it is large, as in Austria and Germany, part of it reflects occupations, and
the gap inside cells sits a few points below the headline without disappearing. Elsewhere the
cells are suppressed for confidentiality, and the analysis does not guess.

## 4. Half the workforce reports every year from 2027

```{r size, fig.height = 3.2}
fig_size(a)
```

Across the EU, `r f$eu_250` of employees in enterprises with ten or more staff work in one of
250 or more, from `r f$min_250_share` in `r f$min_250` to `r f$max_250_share` in `r f$max_250`.
Those employers report every year from June 2027. Employers of 150 to 249 report from 2027 every
three years, and those of 100 to 149 from 2031; the survey groups all enterprises of 50 to 249
together, so the share of workers they cover cannot be separated.

The first wave is also where the gap is widest. In `r f$size_names`, the only countries with
complete figures by size, the gap in enterprises of 250 or more is larger than in smaller ones:
`r f$de_large` against `r f$de_small` in Germany, `r f$pl_large` against `r f$pl_small` in
Poland.

## 5. Where it bites

```{r bites, fig.height = 5.1}
fig_bites(a)
```

Put the two measures together. **`r f$quadrant`** combine a within-sector gap above the EU level
with an above-average share of employees in annual reporters: the first reports will show wide
gaps, and many of them. In `r f$high_within_small`, the gap inside sectors is also above the EU
level, but more workers sit in mid-sized firms. There, the three-yearly reports of employers with
150 to 249 workers, and the right of every worker to ask for pay information (Article 7), will
carry more of the weight.

## 6. Two things the headline does show

```{r control, fig.height = 5.3}
fig_control(a)
```

**The private sector carries the wider gap.** In `r f$public_lower` of the
`r f$public_compared` countries that publish both, the gap is smaller in enterprises under
public control; `r f$public_exception` is the exception. The Directive applies to public
employers too, but the pressure will be greater in private ones.

```{r age, fig.height = 3.6}
fig_age(a)
```

**The gap opens with age.** In the median country it is `r f$age_young` for workers under 25
and peaks at `r f$age_peak` at `r f$age_peak_group`. Gaps build through careers, not starting
salaries, which is why the Directive's requirement to make pay progression criteria accessible
(Article 6) matters as much as the reporting itself.

## 7. What employers should do now

1. **Stop benchmarking against the national gap.** It averages over where women and men work;
   the Directive does not.
2. **Measure inside your own categories of workers,** with a gender-neutral job evaluation, and
   look at progression and at bonuses and allowances as well as base pay.
3. **Run a dry run on 2026 payroll before June 2027.** A worked example of the whole exercise,
   from the seven reporting items to the joint pay assessment screen, is in the
   [Pay Transparency Readiness Kit](https://github.com/D0M3N1C0X/pay-transparency-readiness-kit)
   [@perroni2026kit].
4. **Be most careful where the headline looks reassuring.** `r f$low_but_hidden` publish gaps
   below the EU level that widen by five points or more inside sectors.

## Method in brief

::: {.method}
**Data.** Eurostat's unadjusted gender pay gap series (2010-`r f$year`) and the Structure of
Earnings Survey 2022, downloaded on `r f$retrieved` through the Eurostat API and kept as
snapshots in the repository. Scope as in the official gap: enterprises with ten or more
employees, NACE sections B to S excluding O.

**Decomposition.** For each country, the gap in mean hourly earnings is split into a between-cell
and a within-cell part with symmetric weights [@oaxaca1973; @blinder1973; @reimers1983]:
employment shares and wages of women and men are averaged, so the parts sum to the gap exactly.
A country is used only if the cells cover at least 95% of each sex's employees and rebuild the
published 2022 gap within 0.5 points. All `r f$countries` countries pass with sector cells (the
largest difference is `r f$max_diff` points). The EU figure is the employee-weighted mean of the
national ones, which is how Eurostat compiles it; pooling EU earnings instead would give
`r f$pooled`.

**Limits.** Cells are much coarser than an employer's categories of workers, so the within part
is a guide to what employers will find, not an estimate of discrimination [@blau2017]. The 2022
survey predates the Directive, excludes public administration and enterprises with fewer than
ten employees, and measures enterprises, not legal employers.

The full method, robustness checks and country tables are in the
[working paper](paper.pdf). Data, code and an interactive dashboard:
[github.com/D0M3N1C0X/where-pay-transparency-bites](https://github.com/D0M3N1C0X/where-pay-transparency-bites).
:::

## References

::: {#refs}
:::
