---
title: "Where pay transparency bites"
subtitle: "Sectoral composition, enterprise size and the EU gender pay gap before Directive (EU) 2023/970"
author:
  - name: Domenico Perroni
    orcid: 0009-0001-8806-5188
    affiliations:
      - name: Independent researcher, Kraków
date: 2026-09-17
date-format: "D MMMM YYYY"
abstract: |
  From June 2027, EU employers with 250 or more workers must report their gender pay gap every
  year and assess any gap of 5% or more between women and men doing work of equal value. This
  paper asks how the published national gap relates to the comparisons the Directive requires.
  Using Eurostat's Structure of Earnings Survey 2022, the gap in each of the 27 Member States is
  split, with symmetric weights, into a part due to the sectors women and men work in and a part
  within sectors. Sector cells rebuild every published national gap to within half a point. The
  within-sector gap exceeds the published gap in 22 countries and is 14.6% for the EU against a
  published 12.2%, because women are over-represented in sectors that pay above average. The
  difference is largest where the headline is smallest, as in Italy and Romania. Where earnings
  by enterprise size are published, the gap is wider in enterprises of 250 or more, which employ
  half the workforce covered. Employers benchmarking against the national figure should expect
  their first reports to show wider gaps.
keywords: [gender pay gap, pay transparency, decomposition, Structure of Earnings Survey, Directive (EU) 2023/970]
bibliography: references.bib
format:
  typst:
    papersize: a4
    margin:
      x: 2.4cm
      y: 2.4cm
    fontsize: 10.5pt
    section-numbering: "1.1"
    fig-width: 6.2
    fig-height: 3.8
    keep-typ: false
    citeproc: true
---

```{r setup}
suppressPackageStartupMessages({library(dplyr); library(tidyr); library(ggplot2)})
for (f in list.files("R", pattern = "[.]R$", full.names = TRUE)) source(f)
a <- run_analysis()
f <- facts(a)
```

*Version 1.0. JEL: J16, J31, J71, K31. Data, code and a reproducible build:
github.com/D0M3N1C0X/where-pay-transparency-bites. Every figure in the text and tables is
generated by that code; the test suite checks the abstract against it.*

# Introduction

Directive (EU) 2023/970 [@directive2023] turns the gender pay gap from a national statistic into
an obligation for employers. Those with 250 or more workers report seven indicators every year
from 7 June 2027, those with 150 to 249 every three years from the same date, and those with 100
to 149 from 2031 (Article 9). Where a gap of at least 5% in any category of workers is neither
justified on objective, gender-neutral criteria nor remedied within six months, the employer must
carry out a joint pay assessment with workers' representatives (Article 10).

The benchmark most employers and commentators have at hand is the unadjusted gap Eurostat
publishes for each country. In `r f$year` it was `r f$eu_latest` for the EU, down from
`r f$eu_2010` in 2010. That figure compares all women and all men in the economy. The Directive
compares women and men within one employer and, for the joint pay assessment, within categories
of workers doing the same work or work of equal value.

This paper asks how far apart those two comparisons are likely to be. It splits the published
2022 gap in each Member State into a part due to where women and men work and a part due to how
they are paid within the same sector, and it measures how much of the workforce falls under the
first reporting wave. Decompositions of the gap are well established, including on the same
survey with microdata [@leythienne2018; @blau2017]. The contribution here is narrower and
practical: a fully reproducible decomposition on public tables, reconciled with the published
figures for every country, and read against the Directive's timetable.

# Data

**Sources.** Two Eurostat sources, downloaded through the dissemination API on `r f$retrieved`
and stored as snapshots with the build:

- the unadjusted gender pay gap series (`earn_gr_gpgr2`, with breakdowns by economic control and
  by age), `r min(a$headline$year)` to `r f$year`;
- the Structure of Earnings Survey (SES) 2022: numbers of employees and mean gross hourly
  earnings in euro by sex, economic activity, occupation and enterprise size (`earn_ses22_02`,
  `_13`, `_18`, `_47`, `_53`, `_54`).

**Scope.** The analysis follows the definition of the published gap: enterprises with ten or more
employees, NACE Rev. 2 sections B to S excluding O, mean gross hourly earnings including paid
overtime and excluding non-regular payments [@eurostat_gpg_meta]. Cells are the 17 sections in
that scope; ISCO-08 major groups; or sections crossed with major groups.

**Confidentiality.** Eurostat suppresses cells that could identify respondents. A country enters a
decomposition only if the cells with both a count and a wage cover at least
`r fmt_share(COVERAGE_MIN)` of each sex's employees and rebuild the published 2022 gap within
`r RECONCILE_MAX` percentage points. With sector cells, all `r f$countries` countries pass; with
occupation cells, `r f$finer_occ`; with sector-by-occupation cells, `r f$finer_so`. Earnings by
enterprise size are complete for `r f$size_countries` countries; employee counts by size are
complete for all.

# Method

Let $n_{gc}$ be the number of employees of sex $g \in \{M, F\}$ in cell $c$,
$s_{gc} = n_{gc} / \sum_c n_{gc}$ the employment share, and $w_{gc}$ mean hourly earnings, so that
the mean for each sex is $\bar{W}_g = \sum_c s_{gc} w_{gc}$. The difference in means splits
exactly into

$$
\bar{W}_M - \bar{W}_F
= \underbrace{\sum_c (s_{Mc} - s_{Fc}) \, \frac{w_{Mc} + w_{Fc}}{2}}_{\text{between cells}}
+ \underbrace{\sum_c \frac{s_{Mc} + s_{Fc}}{2} \, (w_{Mc} - w_{Fc})}_{\text{within cells}} .
$$

Dividing by $\bar{W}_M$ expresses both parts in points of the gap as Eurostat defines it. The
averaged weights follow @reimers1983 and avoid the index-number problem of the one-sided split of
@oaxaca1973 and @blinder1973, in which the result depends on whose wages value the differences in
composition. The one-sided splits are reported as a robustness check; the symmetric one is their
mean.

This is an aggregate decomposition: it uses cell means, not individual records, and it attributes
nothing to education, experience or hours. The within-cell part is therefore not an adjusted gap
or an estimate of discrimination. It measures how different the pay of women and men is once they
are compared inside the same broad sector, which is the direction in which the Directive moves
the comparison.

When a cell has no employees of one sex, the other sex's wage is used for both, so the cell adds
nothing within and contributes only through composition. The EU figure is the employee-weighted
mean of the national figures, which is how Eurostat compiles the published EU gap
[@eurostat_gpg_meta].

# Results

## The published gap

```{r trend, fig.height = 3.4}
fig_trend(a)
```

The EU gap fell from `r f$eu_2010` in 2010 to `r f$eu_latest` in `r f$year`, narrowing in
`r f$narrowed` of `r f$compared` countries and widening in `r f$widened`. In `r f$year` it ranged
from `r f$lowest_gap` (`r f$lowest`) to `r f$highest_gap` (`r f$highest`).

## Between and within sectors

```{r hidden, fig.height = 5.8}
fig_hidden(a)
```

For the EU, the published 2022 gap of `r f$eu_2022` splits into `r f$eu_between` between sectors
and `r f$eu_within` within them. The between-sector part is negative in `r f$between_negative`
countries, and the within-sector gap exceeds the published one in `r f$within_above`. The
difference is largest in `r f$hidden_top` (Table A1).

The negative composition effect has a simple source. Women are concentrated in sections with
above-average hourly earnings, above all education (P) and health and social work (Q). In Italy
`r f$it_education` of women in scope work in education against `r f$it_education_m` of men, and
`r f$it_health` in health and social work against `r f$it_health_m`; composition lowers the
Italian gap by `r sub("-", "", f$it$between)`, leaving `r f$it$within` within sectors. A low
headline can also reflect selection into employment, since where fewer women work those who do
tend to have higher-wage characteristics [@olivetti2008]. The decomposition does not separate the
two explanations, but both imply that the national figure understates the gap employers will
measure among their own staff.

## Finer cells

```{r finer}
#| tbl-colwidths: [28, 18, 27, 27]
finer <- a$decomposition |>
  filter(cells != "sector", usable, geo %in% EU27) |>
  mutate(Country = unname(a$names[geo])) |>
  select(Country, cells, official, within) |>
  pivot_wider(names_from = cells, values_from = within) |>
  arrange(Country) |>
  select(Country, official, occupation, `sector x occupation`)
knitr::kable(
  finer |> mutate(across(where(is.numeric), \(x) ifelse(is.na(x), "", formatC(x, format = "f", digits = 1)))),
  col.names = c("Country", "Published gap", "Within occupation", "Within sector and occupation"),
  align = "lrrr",
  caption = "Gap within finer cells, 2022, percentage points. Countries with complete cells only."
)
```

With finer cells the within part stays between `r f$finer_min` and `r f$finer_max`. In
low-headline countries it confirms a wide gap among women and men in similar work; in
high-headline countries such as Austria and Germany, occupations account for part of the gap and
the within part falls a few points below the headline.

## Enterprise size

```{r size, fig.height = 2.9}
fig_size(a)
```

Enterprises of 250 or more employ `r f$eu_250` of employees in scope across the EU, from
`r f$min_250_share` in `r f$min_250` to `r f$max_250_share` in `r f$max_250`; enterprises of 50
to 249 employ a further `r f$eu_50`. SES size classes do not separate the Directive's 100 and 150
thresholds. In all `r f$size_countries` countries with complete earnings by size, the gap is
wider in enterprises of 250 or more.

```{r bites, fig.height = 4.6}
fig_bites(a)
```

`r f$quadrant` combine a within-sector gap above the EU value with an above-average share of
employees in annual reporters. `r f$high_within_small` have wide within-sector gaps but more
employment in mid-sized enterprises, where reporting is three-yearly or later.

## Economic control and age

```{r control, fig.height = 4.8}
fig_control(a)
```

In `r f$public_lower` of `r f$public_compared` countries the gap is smaller under public control
(the exception is `r f$public_exception`). Across countries, the median gap rises from
`r f$age_young` under 25 to `r f$age_peak` at `r f$age_peak_group`.

# Robustness

**Reference wages.** Valuing composition at men's wages, the within-sector gap exceeds the
published one in `r f$within_above_men` countries; at women's wages, in `r f$within_above_women`.
The one-sided within parts differ from the symmetric one by at most `r f$ref_dev`.

**Reconciliation.** Sector cells rebuild every published national gap; the largest difference is
`r f$max_diff` points. The EU figure rebuilt as an employee-weighted mean matches the published
`r f$eu_2022`; pooling EU earnings instead gives `r f$pooled`, because it mixes countries' wage
levels.

**Cell size.** The within part is sensitive to how coarse the cells are, as the finer
decompositions show. Categories of workers inside an employer are much narrower than any cell
here, so the within part should be read as a guide to direction, not to the size of the gaps
employers will report.

# Implications

Three points follow for employers and advisers preparing the first reports. First, the national
gap is a poor benchmark: in most Member States the comparison the Directive requires is closer to
the within-sector view, which is wider. Second, the difference is largest where the headline is
smallest, so employers in `r f$low_but_hidden` have the most reason not to be reassured by it.
Third, the first wave covers half of the workforce in scope and, where it can be measured, the
wider gaps. A dry run on current payroll, as in the Pay Transparency Readiness Kit
[@perroni2026kit], is the practical way to find out before June 2027.

# Limitations

The SES 2022 predates the Directive and excludes public administration (section O), which the
Directive covers, and enterprises with fewer than ten employees. Enterprises are statistical units
and may not coincide with the legal employers that report. The analysis uses published cell means,
so it cannot control for individual characteristics, and suppressed cells restrict the finer
decompositions to a subset of countries. Gaps are in mean hourly earnings; the Directive also asks
for medians, annual pay and complementary or variable components.

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

# Appendix: country results {.unnumbered}

```{r table-a1}
#| tbl-colwidths: [17, 11, 11, 12, 10, 12, 12, 15]
tab <- a$headline_decomposition |>
  filter(geo %in% c(EU27, "EU27_2020")) |>
  left_join(a$exposure |> select(geo, share_250), by = "geo") |>
  left_join(a$latest |> select(geo, latest_gap), by = "geo") |>
  mutate(Country = unname(a$names[geo]), eu = geo == "EU27_2020") |>
  arrange(eu, Country) |>
  transmute(Country, official, gap, between, within,
            coverage = ifelse(is.na(coverage), NA, 100 * coverage), share_250 = 100 * share_250, latest_gap)
knitr::kable(
  tab |> mutate(across(where(is.numeric), \(x) ifelse(is.na(x), "", formatC(x, format = "f", digits = 1)))),
  col.names = c("Country", "Published 2022", "Rebuilt 2022", "Between", "Within", "Coverage %",
                "In 250+ %", paste("Published", f$year)),
  align = "lrrrrrrr",
  caption = "Table A1. Sector decomposition of the 2022 gap, coverage and exposure to annual reporting."
)
```
